Type Error Telemetry at Scale

### 2023-09-23 programming23/paper.tex
related work, drafted

\FILL{} compare \mnonstrict{} and \mstrict{}.
Use edit range~(\cref{t:type-error-survival} for NS vs S)
and error density~(\cref{f:error-delta})

@@

Two further refinements, which reduce the burden on experts to
select points of interest, are adaptive bug isolation~\cite{nl-icse-2010}
and blame-proportional logging~\cite{lnsmc-usenix-2018}.
Adaptive bug isolation starts with a set of predicates, studies which are
most correlated with failures, and experiments with adding telemetry to nearby
predicates; this technique can reduce the overhead of telemetry by two orders
of magnitude relative to naive binary instrumentation.
Blame-proportional logging starts with lightweight telemetry to recognize
defects, assigns ranks to methods estimating their likelihood as root causes,
and uses the ranks and future observations to narrow down the cause.
Deep transfer learning is another promising way to automatically hone in on events of
interest~\cite{zfstt-ieeesensors-2022}.

@@

\paragraph{User Studies: Errors and Type Errors}

Foundational:

Mind your language~\cite{mfk-onward-2011}.
60 students (from 120, self-selected),
50 min per week for 6 weeks,
edit sequences,
followup interviews,
huge problem: vocabulary (function body, arity),
problem: semantics of highlight,

\cite{wk-koli-2020,lgk-pj-2022,mcpcspss-abz-2020}
more edit sequences

Ballpark:

\cite{t-hatra-2021}
plan for a large user-centered study of proof assistants,
events: up, down, to cursor, interrupt, reset.
... not very deep, look at edit sequences, replay for errors

\cite{anna-russo-kennedy-ms-2006}
BitFit, web-based java ide log num hint requests num, result of check answer
requests num compile, num run, no code?! good we have the same constraint
115 bitfit users, 728 sessions

\cite{bgimgm-cse-2016}
user study,
study Java compiler error messages after an intervention, counting num errors,
counting repeats, counting per user
changing errors helps

\cite{dlc-citicse-2014}
enhanced stx error seems ineffectual

\cite{h-dissertiation-2005}
heeren
constraint-based type inference;
general te criteria:
- target audience (novice / expert)
- output format (text / viz)
- interactivity (yes / no)
- heuristics (yes / no)
- primary-or-external (most type errors are easy, some could really use dedicated help)

\cite{sscwj-oopsla-2017}
learning to blame localizing type errors
NATE numerical analysis of type errors
5000 labeled programs
 errors from class, first fix

\cite{sjw-jfp-2018}
type error witness generation
4500 ill-typed student programs
 \pct{85} of time, synthesis works
 \pct{70} of time, locate source

\cite{rsgl-cpp-2020}
user study coq proof assistant
N devs, XXX

\cite{cdhhjklwya-hatra-2020}
evaluating the PLIERS design process (Coblenz etal 202X)
teaching students to use PLIERS; everyone ran a user study!
6 languages
 3 ran usability studies
recommend talk-alouds

\cite{gstf-hatra-2021}
liquid types for java
30 devs user study
 zoom

\cite{lfgc-pldi-2007}
typechecker never errors,
 calls a helper that looks for nearby programs that do typecheck
tested on
 10 students, 2yrs professional experience
 5 hw assignments
 2122 files collected;
won \pct{19}, lost \pct{17};
ill-suited for things like (let x = e1 in ....),
 change e1,
 use typechecker feedback to guide edits;
small data!

\cite{w-popl-1986}
reporting source, not detection
tested on 9 type errors
 all \emph{deletions}, such as forgetting to inject in an enum

\cite{hw-scp-2004}
haack wells
location of type error as a slice,
 not a srcloc,
 not a subtree

bibliography of static type studies:
http://danluu.com/empirical-pl/

\paragraph{TBD}

thomas kennedy .... generic usage monitoring ascilite 2003

ahmadzadeh elliman higgins iticse 2005

buffardi etal 2014 adaptive and social mechanisms

bddf-icse-2016
 sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform

murphy etal sigcse 2009 Retina: helping students

Choppella Haynes IU TR 426

Tip Dinesh tosem 2001

@@

@@

Timestamps are not the focus of our analysis, but they do play a critical role.
Consider~\cref{f:error-delta}.
With timestamps, it is clear that error density stays close to zero.
Without timestamps, we would be left with the counts at the bottom
of~\cref{f:error-delta}.
Many distributions could fit those counts, such as a ``mountain'' session that first
grows and grows the number of type errors then shrinks and shrinks down to zero.

>>>>

\FILL{} compare \mnonstrict{} and \mstrict{}.
Use edit range~(\cref{t:type-error-survival} for NS vs S)
and error density~(\cref{f:error-density})

@@

Two further refinements, which reduce the burden on experts to
select points of interest, are adaptive bug isolation~\cite{nl-icse-2010}
and blame-proportional logging~\cite{lnsmc-usenix-2018}.
Adaptive bug isolation starts with a set of predicates, studies which are
most correlated with failures, and experiments with adding telemetry to nearby
predicates; this technique can reduce the overhead of telemetry by two orders
of magnitude relative to naive binary instrumentation.
Blame-proportional logging starts with lightweight telemetry to recognize
defects, assigns ranks to methods estimating their likelihood as root causes,
and uses the ranks and future observations to narrow down the cause.
Deep transfer learning is another promising way to hone in on events of
interest~\cite{zfstt-ieeesensors-2022}.

@@

\paragraph{User Studies, Errors, and Type Errors}

Our approach to data analysis is inspired by Marceau
et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011}, who used keystroke-level data
collected over 6 weeks to analyze students' responses to runtime errors
in DrRacket~\cite{fcffksf-jfp-2002}.
Similarly, Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021} capture
edit sessions from the online Alloy4Fun editor and present several automated
analyses.
A major challenge in our setting, however, is that edit sequences are
incomplete---there may be minutes-long gaps between two telemetry records.
Several more-recent works have studied edit sequences
to draw conclusions about errors and intent, e.g.,~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.

The BitFit project collects data about beginning students interactions with a
Java IDE~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; however, unlike Blackbox, it does
not collect source code.
BitFit data consists of a timestamp, a session id, and counter values showing
the number of attempts at a question, the number of hints requested, the number
of compiles and runs, and the number of submission attempts.
On one hand, this data is more informative than Roblox Studio's because it
targets specific events rather than random keystrokes.
On the other hand, there are no details about errors to show what students
struggled with.
Other logging-based studies of Java programs use the full error message to
draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}.
Roblox Studio cannot follow this approach because strings in errors may leak
personal information.

Much work on the benefits of static types focuses on a small number
of case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
or static corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}
(for literature reviews, see~\cite{empirical-types,heeren-thesis}).
Controlled experiments, for example comparing statically-typed Java and dynamically-typed
Groovy~\cite{khrts-icpc-2012}, are a notable exception.
Lerner et~al.~\cite{lfgc-pldi-2007} apply type error repair tool on over two thousand
of programs written during five homework exercises.
Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
to train data-driven methods for localizing errors.
These large studies with several thousands of programs are
nevertheless on a much smaller scale than Roblox Studio.

%\paragraph{TBD}
%
%thomas kennedy .... generic usage monitoring ascilite 2003
%
%ahmadzadeh elliman higgins iticse 2005
%
%buffardi etal 2014 adaptive and social mechanisms
%
%bddf-icse-2016
% sh edwards and jason snyder icer 2009 comparing effective and ineffective grading platform
%
%murphy etal sigcse 2009 Retina: helping students
%
%Choppella Haynes IU TR 426

@@

% \cite{}
% heeren
% constraint-based type inference;
% general te criteria:
% - target audience (novice / expert)
% - output format (text / viz)
% - interactivity (yes / no)
% - heuristics (yes / no)
% - primary-or-external (most type errors are easy, some could really use dedicated help)

% \cite{t-hatra-2021}
% plan for a large user-centered study of proof assistants,
% events: up, down, to cursor, interrupt, reset.
% ... not very deep, look at edit sequences, replay for errors

@@

Timestamps are not the focus of our analysis, but they do play a critical role.
Consider~\cref{f:error-density}.
With timestamps, it is clear that error density stays close to zero.
Without timestamps, we would be left with the counts at the bottom
of~\cref{f:error-density}.
Many distributions could fit those counts, such as a ``mountain'' session that first
grows and grows the number of type errors then shrinks and shrinks down to zero.
